---
title: "Capital One Data Challenge"
author: "Manroop Singh"
date: "September 12, 2016"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r,echo=FALSE, warning=FALSE, message= FALSE}
library(readr)
library(magrittr)
library(dplyr)
library(data.table)
# library(plyr) 
library(dplyr) 
library(ggplot2) 
library(gplots)
library(lubridate)
library(devtools)
library(Rcpp)
#install_github('arilamstein/choroplethrZip@v1.5.0')
library(glmnet)
library(e1071)
library(tm)
library(plotly)
library(manipulate)
```

##About the Data

####Report is based on two datasets

  *Loans Dataset*  
     Details about the applications over the span of three years. 
         
  *Institutions Dataset*  
     Details about the lenders for those loans.  
  


#### Trend among approved loan applications over the years.

```{r, echo=FALSE}

df_loan_raw<-read.csv(file="2012_to_2014_loans_data.csv")
df_institutions_raw<-read.csv(file="2012_to_2014_institutions_data.csv")

df_loan_subset<-data.frame(df_loan_raw$As_of_Year, df_loan_raw$Loan_Purpose_Description)
temp<-as.data.frame.matrix(table(df_loan_subset))
temp<-add_rownames(temp,"Year")
plot(x=temp$Year, xlab = "Year",y=temp$Refinance, ylab="No. of Applications", type="l", col="Red")
lines(x=temp$Year, y=temp$Purchase, col="Green")
title("Loans")
legend('topright', c("Refinance","Purchase"), lty=1, col = c("Red","Green"), cex=0.75)

```

> The number of refinance applications have fallen down over the period



#Demographics as per state
> The dataset have 5 States

####Average of loan amount as per the states


```{r, echo=FALSE}
df_loans_2012<-subset(df_loan_raw, df_loan_raw$As_of_Year == "2012")
df_loans_2013<-subset(df_loan_raw, df_loan_raw$As_of_Year == "2013")
df_loans_2014<-subset(df_loan_raw, df_loan_raw$As_of_Year == "2014")


par(mfrow=c(1,3))
par(mar = c(2,2,2,0))
df_state_avg_loan_2012<-tapply(df_loans_2012$Loan_Amount_000, list(df_loans_2012$State,df_loans_2012$Loan_Purpose_Description), mean)
y_2012<-df_state_avg_loan_2012
barplot(y_2012, main="2012", ylim = c(0,500), col=rainbow(5),beside=TRUE, ylab="Loan Amount (thousands)")

df_state_avg_loan_2013<-tapply(df_loans_2013$Loan_Amount_000, list(df_loans_2013$State,df_loans_2013$Loan_Purpose_Description), mean)
y_2013<-df_state_avg_loan_2013
barplot(y_2013, main="2013", ylim = c(0,500), col=rainbow(5),beside=TRUE, ylab="Loan Amount (thousands)")

df_state_avg_loan_2014<-tapply(df_loans_2014$Loan_Amount_000, list(df_loans_2014$State,df_loans_2014$Loan_Purpose_Description), mean)
y_2014<-df_state_avg_loan_2014
barplot(y_2014, main=" 2014", ylim = c(0,500), col=rainbow(5),beside=TRUE, ylab="Loan Amount (thousands)")
legend("topright", legend=rownames(y_2014), fill=2:6,cex = 0.75)
```

> The average loan amount in DC is way higher than other states.   

####Loans by each agency

Number of loans by each agency over the span of three years
```{r, echo=FALSE}

par(mfrow=c(2,2))
df<-as.data.frame.matrix(df_loan_raw)
loans_by_agency<-table(df$Agency_Code_Description)
marks<-c(0,100000,200000,300000,400000,500000,600000)
barplot(loans_by_agency, yaxt="n", col="light blue", ylim = c(0,600000)) 
axis(2,at=marks,labels=format(marks,scientific=FALSE), las=2)
title("2012-2014")

df_apps_by_agency_2012<-as.data.frame.matrix(df_loans_2012)
loans_by_agency_2012<-table(df_apps_by_agency_2012$Agency_Code_Description)
marks<-c(0,50000, 100000,150000,200000,250000,300000)
barplot(loans_by_agency_2012, yaxt="n", col="light blue", ylim = c(0,300000)) 
axis(2,at=marks,labels=format(marks,scientific=FALSE), las=2)
title("2012")


df_apps_by_agency_2013<-as.data.frame.matrix(df_loans_2013)
loans_by_agency_2013<-table(df_apps_by_agency_2013$Agency_Code_Description)
marks<-c(0,50000, 100000,150000,200000)
barplot(loans_by_agency_2013, yaxt="n", col="light blue", ylim = c(0,200000)) 
axis(2,at=marks,labels=format(marks,scientific=FALSE), las=2)
title("2013")


df_apps_by_agency_2014<-as.data.frame.matrix(df_loans_2014)
loans_by_agency_2014<-table(df_apps_by_agency_2014$Agency_Code_Description)
marks<-c(0,50000, 100000,150000)
barplot(loans_by_agency_2014, yaxt="n", col="light blue", ylim = c(0,150000)) 
axis(2,at=marks,labels=format(marks,scientific=FALSE), las=2)
title("2014")


```


> Clearly the majority of loans are for CFPB and HUD.



#Fair lending as per income

####Income classes have been defined as per following

Upper Class(UC): > *$150,000*  
Upper middle class(UMC): *$90,000-$150,000*  
Lower Middle Class(LMC):*$32,500 - 90,000*  
Below Povery line(BPL): *$32,500*  


#### Number of applicants in each class 


```{r, echo=FALSE}

df_income_class<-df_loan_raw
df_income_class<-df_income_class[complete.cases(df_income_class),]
df_income_class$IncomeClass<-"0"
df_income_class$Applicant_Income_000<-as.numeric(df_income_class$Applicant_Income_000)
df_income_class$IncomeClass<-ifelse(df_income_class$Applicant_Income_000 <= 32.5, "BPL",ifelse(df_income_class$Applicant_Income_000<=90,"LMC",ifelse(df_income_class$Applicant_Income_000<=150, "UMC","UC")))
df_2012_income_class<-subset(df_income_class, df_income_class$As_of_Year == "2012")
df_2013_income_class<-subset(df_income_class, df_income_class$As_of_Year == "2013")
df_2014_income_class<-subset(df_income_class, df_income_class$As_of_Year == "2014")

apps_per_income_class<- table(df_income_class$IncomeClass)

#par(mfrow=c(1,3))
apps_per_income_class_2012<- table(df_2012_income_class$IncomeClass)
#barplot(apps_per_income_class_2012, col = "light blue", ylim = c())
#title("2012")

apps_per_income_class_2013<- table(df_2013_income_class$IncomeClass)
#barplot(apps_per_income_class_2013)
#title("2013")

apps_per_income_class_2014<- table(df_2014_income_class$IncomeClass)
#barplot(apps_per_income_class_2014)
#title("2014")
```


```{r, echo=FALSE,warning=FALSE, message= FALSE}

library(gridExtra)

x2012<-as.data.frame(apps_per_income_class_2012)
plot_x2012<- ggplot(x2012, aes(x = factor(x2012$Var1), y = x2012$Freq,fill =factor(x2012$Var1))) + geom_bar(stat = "identity") + xlab("") + ylab("No of applications") + guides(fill=guide_legend(title=NULL)) + guides(fill=FALSE)


x2013<-as.data.frame(apps_per_income_class_2013)
plot_x2013<-ggplot(x2013, aes(x = factor(x2013$Var1), y = x2013$Freq,fill =factor(x2013$Var1))) + geom_bar(stat = "identity") + xlab("Income Classes") + ylab("") + guides(fill=guide_legend(title=NULL)) + guides(fill=FALSE)


x2014<-as.data.frame(apps_per_income_class_2014)
plot_x2014<-ggplot(x2014, aes(x = factor(x2014$Var1), y = x2014$Freq,fill =factor(x2014$Var1))) + geom_bar(stat = "identity") + xlab("") + ylab("") + guides(fill=guide_legend(title=NULL)) + guides(fill=FALSE)

grid.arrange(plot_x2012, plot_x2013,plot_x2014, ncol=3)

```

#Avg loan as per income

#### For the span for 3 years  
```{r, echo=FALSE}

avg_loans_per_income<-tapply(df_income_class$Loan_Amount_000, df_income_class$IncomeClass, mean)
barplot(avg_loans_per_income, col="light blue", ylab="Avg loan amount(in thousands)", ylim=c(0,500))
title("2012-2014")

```

#### Per year

```{r, echo=FALSE}

par(mfrow=c(1,3))

avg_loans_per_income_2012<-tapply(df_2012_income_class$Loan_Amount_000, df_2012_income_class$IncomeClass, mean)
barplot(avg_loans_per_income, col="light blue", ylab="Avg loan amount(in thousands)", 
ylim=c(0,600))
title("2012")

avg_loans_per_income_2013<-tapply(df_2013_income_class$Loan_Amount_000, df_2013_income_class$IncomeClass, mean)
barplot(avg_loans_per_income_2013, col="light blue", ylab="Avg loan amount(in thousands)", ylim=c(0,600))
title("2013")

avg_loans_per_income_2014<-tapply(df_2014_income_class$Loan_Amount_000, df_2014_income_class$IncomeClass, mean)
barplot(avg_loans_per_income_2014, col="light blue", ylab="Avg loan amount(in thousands)", ylim=c(0,600))
title("2014")


```


#Percentage of loans per income 


```{r, echo=FALSE}
total_loans_per_income_2012<-tapply(df_2012_income_class$Loan_Amount_000, df_2012_income_class$IncomeClass, "sum")
s<-sum(total_loans_per_income_2012)
percent<-(total_loans_per_income_2012/s) * 100
barplot(percent, col="light blue", ylim=c(0,50), ylab = "Share (%)")
title("2012")
```


```{r, echo=FALSE}
total_loans_per_income_2013<-tapply(df_2013_income_class$Loan_Amount_000, df_2013_income_class$IncomeClass, "sum")
s<-sum(total_loans_per_income_2013)
percent<-(total_loans_per_income_2013/s) * 100
barplot(percent, col="light blue", ylim=c(0,50), ylab = "Share (%)")
title("2013")
```


```{r, echo=FALSE}
total_loans_per_income_2014<-tapply(df_2014_income_class$Loan_Amount_000, df_2014_income_class$IncomeClass, "sum")
s<-sum(total_loans_per_income_2014)
percent<-(total_loans_per_income_2014/s) * 100
barplot(percent, col="light blue", ylim=c(0,50), ylab = "Share (%)")
title("2014")
```


#Risk factor analysis by lenders

> This is the percentage of loans approved by the financial institutions in each state, which were approved on Subordinate Lien i.e higher risk.  

```{r, echo=FALSE}

 df_1<-df_loan_raw
df_1$ID = paste0(df_1$Agency_Code,df$Respondent_ID)
df_2<-df_institutions_raw
df_2$ID = paste0(df_2$Agency_Code,df_2$Respondent_ID)
df_merge<-merge(x=df_1, y=df_2, by = "ID")
lien_status_statewise<-table(df_merge$Respondent_State_TS, df_merge$Lien_Status_Description)
lien_status_statewise<-as.data.frame.matrix(lien_status_statewise)
lien_status_statewise$sum<- lien_status_statewise$`First Lien` + lien_status_statewise$`Subordinate Lien`
lien_status_statewise$risk_factor<-(lien_status_statewise$`Subordinate Lien`/lien_status_statewise$sum)*100
lien_status_statewise$risk_factor<-round(lien_status_statewise$risk_factor, digits=2)
lien_status_statewise<-lien_status_statewise[complete.cases(lien_status_statewise), ]
lien_status_statewise<-lien_status_statewise[order(-lien_status_statewise$risk_factor), ]
top_risk_taker<-lien_status_statewise[1:10,]
top_risk_taker<-add_rownames(top_risk_taker,"State")
barplot(top_risk_taker$risk_factor,names.arg = top_risk_taker$State, ylab="% of loans on Subordinate Lien", col=terrain.colors(10))

```

> As shown in the rate of highest risk taker in decreasing order










